---
title: "MRFs and CRFs for Bird.parasites data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MRFs and CRFs for Bird.parasites data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

We can explore the model's primary functions using a test dataset that is available with the package. Load the `Bird.parasites` dataset, which contains binary occurrences of four avian blood parasites in New Caledonian *Zosterops* species ([available in its original form at Dryad](http://dx.doi.org/10.5061/dryad.pp6k4); Clark *et al* 2016). A single continuous covariate is also included (`scale.prop.zos`), which reflects the relative abundance of *Zosterops* species (compared to other avian species) among different sample sites
```{r, message=FALSE, warning=FALSE}
library(MRFcov)
data("Bird.parasites")
```

The `Bird.parasites` dataset is already in the appropriate structure for running `MRFcov` models; however, it is useful to understand how this was done. Using the raw dataset (downloaded from `Dryad` at the above link), we created a scaled continuous covariate to represent *Zosterops* spp. proportional captures in each sample site (as an estimate of relative abundance)
```{r,message=F, warning=FALSE, eval = FALSE}
#Not run
#install.packages(dplyr)
data.paras = data.frame(data.paras) %>%
  dplyr::group_by(Capturesession,Genus) %>%
  dplyr::summarise(count = dlyr::n()) %>%
  dplyr::mutate(prop.zos = count / sum(count)) %>%
  dplyr::left_join(data.paras) %>%
  dplyr::ungroup() %>% dplyr::filter(Genus == 'Zosterops') %>%
  dplyr::mutate(scale.prop.zos = as.vector(scale(prop.zos)))
data.paras <- data.paras[, c(12:15, 23)]
```

You can visualise the dataset to see how analysis data needs to be structured. In short, for (`family = "binomial"`) models, node variable (i.e. species) occurrences should be included as binary variables (1s and 0s) as the left-most variables in `data`. Note that Gaussian continuous variables (`family = "gaussian"`) and Poisson non-negative counts (`family = "poisson"`) are also supported in `MRFcov`. Any covariates can be included as the right-most variables. It is recommended that these covariates all be on a similar scale, ideally using the `scale` function for continuous covariates (or similar) so that covariates have roughly `mean = 0` and `sd = 1`, as this makes LASSO variable selection and comparisons of effect sizes very straightforward
```{r eval=FALSE}
help("Bird.parasites")
View(Bird.parasites)
```

### Running MRFs and visualising interaction coefficients
We first test for interactions using a Markov Random Fields (MRF) model without covariates. We set `n_nodes` as the number of species (`4`). We also need to specify the family for the model (`binomial` in this case). We have two options for specifying the level of penalisation used by the LASSO algorithm. We can either specify a single `lambda1` value, which results in fitting penalised regressions using the same `lambda1` for each species (using function `penalized` in the `penalized` package), or we can allow `lambda1` to be optimised separately for each species through cross validation (using function `cv.glmnet` in the `glmnet` package). It is generally advisable to use the second option (which is the default if `lambda1` is not supplied), as this ensures the log-likelihood of each species is properly maximised before unifying them into an undirected graph. Finally, it is important to note that for MRF model functions, we can specify the number of processing cores to split the job across (i.e. `n_cores`). In this case we use 3 cores, but higher numbers *may* increase speed (if more cores are available). Check available cores using the `detectCores()` function in the `parallel` package
```{r,eval=FALSE}
MRF_fit <- MRFcov(data = Bird.parasites[, c(1:4)], n_nodes = 4, family = 'binomial')
```

Plot the estimated interaction coefficients as a heatmap using `plotMRF_hm`. Note that we can specify the names of nodes should we want to change them
```{r eval=FALSE}
plotMRF_hm(MRF_mod = MRF_fit, main = 'MRF (no covariates)', 
                             node_names = c('H. zosteropis', 'H. killangoi',
                                            'Plasmodium', 'Microfilaria'))
```

As a check of the model's estimation accuracy, we can run the same model using functions in David Harris' `rosalia` package, which directly solves small (i.e. less than 20 species, with no covariates) MRF networks by maximum likelihood. We specify a weakly informative logistic prior to ensure finite parameter estimates (following Harris, 2016)
```{r eval=FALSE}
library(rosalia)
rosalia_fit <- rosalia(Bird.parasites[,1:4], 
                      prior = make_logistic_prior(scale = 2),
                      trace = FALSE)
plotMRF_hm(MRF_mod = rosalia_fit, node_names = c('H. zosteropis', 'H. killangoi',
                                           'Plasmodium', 'Microfilaria'),
           main = 'rosalia interactions')
```

We can now calculate proportions of positive and negative interactions found by `rosalia` that were also found by our model. Values approaching 1 suggest strong concordance between the two models
```{r eval=FALSE}
comp_rosalia_MRF(MRF_mod = MRF_fit, rosalia_mod = rosalia_fit)
```

### Running CRFs using additional covariates
We can now run a Conditional Random Fields (CRF) model using the provided continuous covariate (`scale.prop.zos`). Again, we use the default cross-validation method to optimise each species' regression. Note that any columns in `data` to the right of column `n_nodes` will be presumed to represent covariates if we don't specify an `n_covariates` argument
```{r eval=FALSE}
MRF_mod <- MRFcov(data = Bird.parasites, n_nodes = 4, family = 'binomial')
```

Visualise the estimated species interaction coefficients as a heatmap. These represent predicted interactions when the covariate is set to its mean (i.e. in this case, when `scale.prop.zos` = 0)
```{r eval=FALSE}
plotMRF_hm(MRF_mod = MRF_mod)
```

We can also visualise how species interactions are predicted to change across covariate magnitudes. Because we are using a continuous covariate, we can use the function `plotMRF_HM_cont` to visualise changing interactions (`cont` stands for continuous, in this cse)
```{r eval=FALSE}
plotMRF_hm_cont(MRF_mod = MRF_mod, covariate = 'scale.prop.zos', data = Bird.parasites, 
                main = 'Estimated interactions across host relative densities')
```

Regression coefficients and their relative importances can be accessed as well. This call returns a matrix of the raw coefficient, as well as standardised coefficients (standardised by the `sd` of the covariate). Standardisation in this way helps to compare the relative influences of each parameter on the target species' occurrence probability. The list also contains contains each variable's relative importance (calculated using the formula `B^2 / sum(B^2)`, where the vector of `B`s represents regression coefficients for predictor variables). Variables with an underscore (`_`) indicate an interaction between a covariate and another node, suggesting that conditional dependencies of the two nodes vary across environmental gradients
```{r eval = FALSE}
MRF_mod$key_coefs$Hzosteropis
```

Finally, a useful capability is to generate some fake data and test predictions. For instance, say we want to know how frequently malaria parasite infections are predicted to occur in sites with high occurrence of microfilaria
```{r eval = FALSE}
fake.dat <- Bird.parasites
fake.dat$Microfilaria <- rbinom(nrow(Bird.parasites), 1, 0.8)
fake.preds <- predict_MRF(data = fake.dat, MRF_mod = MRF_mod)
```

The returned object from `predict_MRF` depends on the family of the model. For `family = "binomial"`, we get a list including both linear prediction probabilities and binary predictions (where a linear prediction probability `> 0.5` equates to a binary prediction of `1`). These binary predictions can be used to estimate parasite prevalence
```{r eval = FALSE}
H.zos.pred.prev <- sum(fake.preds$Binary_predictions[, 'Hzosteropis']) / nrow(fake.preds$Binary_predictions)
Plas.pred.prev <- sum(fake.preds$Binary_predictions[, 'Plas']) / nrow(fake.preds$Binary_predictions)
Plas.pred.prev
```

### Comparing fits of MRF and CRF models
A key step in the process of model exploration is to determine whether inclusion of covariates actually improves model fit and is warranted when estimating interaction parameters. This is straightforward for binomial models, as we can compare classification accuracy quickly and easily. The `cv_diag` functions will fit models and then determine their predictive performances against the supplied data. We can also compare MRF and CRF models directly to determine whether covariates are warranted. For this dataset, considering that parasite infections are quite rare, we are primarily interested in maximising model Sensitivity (capacity to successfully predict positive infections)
```{r eval = FALSE}
cv_MRF_diag_rep(data = Bird.parasites, n_nodes = 4,
            n_cores = 3, family = 'binomial', plot = T, compare_null = T)
```

### Bootstrapping the data and running models
We now may want to fit models to bootstrapped subsets of the data to account for parameter uncertainty
```{r eval=FALSE}
booted_MRF <- bootstrap_MRF(data = Bird.parasites, n_nodes = 4, family = 'binomial', n_bootstraps = 25, n_cores = 3)
```

### Exploring regression coefficients and interpreting results
Finally, we can explore regression coefficients to get a better understanding of just how important interactions are for predicting species' occurrence probabilities (in comparison to other covariates). This is perhaps the strongest property of CRFs, as comparing the relative importances of interactions and fixed covariates using competing methods (such as Joint Species Distribution Models) is difficult. The `bootstrap_MRF` function conveniently returns a matrix of important coefficients for each node in the graph, as well as their relative importances
```{r eval=FALSE}
booted_MRF$mean_key_coefs$Hzosteropis
```

```{r eval=FALSE}
booted_MRF$mean_key_coefs$Hkillangoi
```

```{r eval=FALSE}
booted_MRF$mean_key_coefs$Plas
```

```{r eval=FALSE}
booted_MRF$mean_key_coefs$Microfilaria
```

### Accounting for possible spatial autocorrelation
Lastly, `MRFcov` has the capability to account for possible spatial autocorrelation when estimating interaction parameters. To do this, we incorporate functions from the `mgcv` package to include smoothed Gaussian process spatial regression splines in each node-wise regression. The user must supply a two-column `data.frame` called `coords`, which will ideally contain Latitude and Longitude values for each row in `data`. We don't have these coordinates for the `Bird.parasites` dataset, so we will instead create some fake coordinates to showcase the model
```{r eval = FALSE}
Latitude <- sample(seq(120, 140, length.out = 100), nrow(Bird.parasites), TRUE)
Longitude <- sample(seq(-19, -22, length.out = 100), nrow(Bird.parasites), TRUE)
coords <- data.frame(Latitude = Latitude, Longitude = Longitude)
```

The syntax for the `MRFcov_spatial` function is nearly identical to `MRFcov`, with the exception that `coords` must be supplied
```{r eval = FALSE}
CRFmod_spatial <- MRFcov_spatial(data = Bird.parasites, n_nodes = 4,
                                 family = 'binomial', coords = coords)
```

Interpretation is also identical to `MRFcov` objects. Here, key coefficients are those that are retained *after* accounting for spatial influences
```{r eval = FALSE}
CRFmod_spatial$key_coefs$Hzosteropis
```

Finally, we can compare fits of spatial and non-spatial models just as we did for MRFs and CRFs above
```{r eval = FALSE}
cv_MRF_diag_rep_spatial(data = Bird.parasites, n_nodes = 4,
                        n_cores = 3, family = 'binomial', plot = T, compare_null = T,
                        coords = coords)
```


